---
title: "Introduction to Forecasting"
subtitle: "Predicting Salmon Returns"  
author: 
#  - "Mark Scheuerell"
  - "WA-BC AFS Meeting<br>Spokane, WA<br><br>29 April 2024"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "my-theme.css"]
    nature:
      slideNumberFormat: ""
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      # countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=8, fig.height=4, fig.retina=3, fig.align = "center",
  out.width = "100%",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

## set random seed
set.seed(666)

## load {icons} for font awesome
library(icons)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo(
  primary_color = "#FFFFFF",
  secondary_color = "#23395b",
  # secondary_color = "#FF961C",
  title_slide_text_color = "#FFFFFF",
  title_slide_background_color = "#23395b",
  colors = c(
    red = "#f34213",
    purple = "#844870",
    orange = "#FF961C",
    green = "#339933",
    white = "#FFFFFF",
    blue = "#488fdf",
    aqua = "#80cdc1",
    gray = "#808080",
    lightgray = "#bdbdbd"
),
  header_font_google = google_font("Roboto Condensed", "400", "400i", "700"),
  text_font_google = google_font("Cabin", "400", "400i", "700"),
)
```


# Forecasting strategy

.pull-left[
### .blue[1) Assess & understand objectives]
]


---

# 1) Assess & understand objectives

### .blue[Who is involved?]

--

### .blue[Harvest vs conservation?]

--

### .blue[Escapement goal vs exploitation rate?]

--

### .blue[Pre-season vs in-season?]


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .blue[2) Gather information]
]


---

# 2) Gather information

### .blue[Who holds the data?]

--

### .blue[Harvest + escapement?]

--

### .blue[Age composition?]

--

### .blue[Smolts?]

--

### .blue[Environmental covariates?]


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .green[2) Gather information]

### .blue[3) Decide on modeling framework]
]

---

# 3) Decide on modeling framework

### .blue[Sibling regression vs environmental? Both?]

--

### .blue[Observation & process errors? (state-space)]

--

### .blue[Counts? Age comp? Mark-recapture? (integrated population model)]


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .green[2) Gather information]

### .green[3) Decide on modeling framework]

### .blue[4) Choose time horizons]
]

---

# 4a) Choose time horizons | fitting vs forecasting

```{r forecast_horizon, out.width = "90%", out.height = "90%"}

# sz <- function(mu, var) {
#   mu^2 / (var - mu)
# }
# 
# sz(1000, 100000) ~= 10

## data
TT <- 30
tt <- seq(TT)
dd <- rnbinom(TT, mu = 15, size = 30)

## plot
par(mai = c(1, 1, 0.1, 1), omi = rep(0, 4))

plot(
  tt, dd,
  type = "o", pch = 19,
  cex = 1.5, cex.axis = 1.2, cex.lab = 1.5,
  las = 1, bty = "n",
  ylab = "Total returns (1000s)", xlab = "Time",
  lwd = 2, col = "dodgerblue"
)
arrows(
  x0 = 30, y0 = -2, x1 = 30, y1 = 0.5,
  length = 0.1, angle = 25, code = 2,
  col = "darkred", lwd = 2,
  xpd = NA
)
text(
  x = 30, y = -2, pos = 1,
  labels = "We are here", cex = 1.5,
  col = "darkred",
  xpd = NA
)
```


---

class: center

# Balanced but perhaps overly optimistic

```{r forecast_1515, out.width = "90%", out.height = "90%"}

## plot
par(mai = c(1, 1, 0.1, 1), omi = rep(0, 4))

plot(
  tt, dd,
  type = "n", cex.axis = 1.2, cex.lab = 1.5,
  las = 1, bty = "n",
  ylab = "Total returns (1000s)", xlab = "Time"
)
## fitting
lines(
  tt[1:15], dd[1:15],
  type = "o", pch = 19, cex = 1.5, lwd = 2,
  col = "dodgerblue"
)
text(
  x = 8, y = 25, pos = 1,
  labels = "Fitting", cex = 1.5,
  col = "dodgerblue",
  xpd = NA
)
## prediction
lines(
  tt[16:TT], dd[16:TT],
  type = "o", pch = 21, cex = 1.5, lwd = 2,
  col = "dodgerblue", bg = "white"
)
text(
  x = 23, y = 0, pos = 1,
  labels = "Prediction", cex = 1.5,
  col = "dodgerblue",
  xpd = NA
)
```


---

class: center

# Fitting-focused but evaluation challenged

```{r forecast_2505, out.width = "90%", out.height = "90%"}

## plot
par(mai = c(1, 1, 0.1, 1), omi = rep(0, 4))

plot(
  tt, dd,
  type = "n", cex.axis = 1.2, cex.lab = 1.5,
  las = 1, bty = "n",
  ylab = "Total returns (1000s)", xlab = "Time"
)
## fitting
lines(
  tt[1:25], dd[1:25],
  type = "o", pch = 19, cex = 1.5, lwd = 2,
  col = "dodgerblue"
)
text(
  x = 13, y = 25, pos = 1,
  labels = "Fitting", cex = 1.5,
  col = "dodgerblue",
  xpd = NA
)
## prediction
lines(
  tt[26:TT], dd[26:TT],
  type = "o", pch = 21, cex = 1.5, lwd = 2,
  col = "dodgerblue", bg = "white"
)
text(
  x = 28, y = 0, pos = 1,
  labels = "Prediction", cex = 1.5,
  col = "dodgerblue",
  xpd = NA
)
```


---

# 4b) Choose time horizons | steps ahead?

```{r forecast_ahead, out.width = "90%", out.height = "90%"}

## plot
par(mai = c(1, 1, 0.1, 1), omi = rep(0, 4))

plot(
  tt, dd,
  type = "n", cex.axis = 1.2, cex.lab = 1.5,
  las = 1, bty = "n",
  ylab = "Total returns (1000s)", xlab = "Time"
)
## fitting
lines(
  tt[1:25], dd[1:25],
  type = "o", pch = 19, cex = 1.5, lwd = 2,
  col = "dodgerblue"
)
text(
  x = 13, y = 25, pos = 1,
  labels = "Fitting", cex = 1.5,
  col = "dodgerblue",
  xpd = NA
)
arrows(
  x0 = 25, y0 = -2, x1 = 25, y1 = 0.5,
  length = 0.1, angle = 25, code = 2,
  col = "darkred", lwd = 2,
  xpd = NA
)
text(
  x = 25, y = -2, pos = 1,
  labels = "We are here", cex = 1.5,
  col = "darkred",
  xpd = NA
)
## prediction
points(
  tt[26:TT], rep(17, 5),
  pch = "?", cex = 1.5, lwd = 2,
  col = "dodgerblue", bg = "white"
)
# text(
#   x = 28, y = 0, pos = 1,
#   labels = "Prediction", cex = 1.5,
#   col = "dodgerblue",
#   xpd = NA
# )
```


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .green[2) Gather information]

### .green[3) Decide on modeling framework]

### .green[4) Choose time horizons]
]

.pull-right[
### .blue[5) Fit model & generate forecast]
]


---

# 5) Fit model & generate forecast

### .blue[Bayesian? Priors?]

--

### .blue[Software?]


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .green[2) Gather information]

### .green[3) Decide on modeling framework]

### .green[4) Choose time horizons]
]

.pull-right[
### .green[5) Fit model & generate forecast]

### .blue[6) Evaluate fit & skill]
]


---

# 6) Evaluate fit & skill

### .blue[Accuracy & Precision?]

--

### .blue[Model comparisons? Versus naïve?]


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .green[2) Gather information]

### .green[3) Decide on modeling framework]

### .green[4) Choose time horizons]
]

.pull-right[
### .green[5) Fit model & generate forecast]

### .green[6) Evaluate fit & skill]

### .blue[7) Report results]

]


---

# 7) Report results

### .blue[Clearly communicated?]

--

### .blue[Uncertainties?]

--

### .blue[Open and reproducible?]


---

# Forecasting strategy

.pull-left[
### .green[1) Assess & understand objectives]

### .green[2) Gather information]

### .green[3) Decide on modeling framework]

### .green[4) Choose time horizons]
]

.pull-right[
### .green[5) Fit model & generate forecast]

### .green[6) Evaluate fit & skill]

### .green[7) Report results]

### .blue[8) Refine & repeat]
]


---

# 8) Refine & repeat

### .blue[How did we do? Were mistakes made?]

--

### .blue[Update data & priors?]

--

### .blue[Update model(s)?]


